How to write cross-interpreter programs

### 2014-06-05 talk/dls2014/paper/paper.tex
Attempt to simplify the Memory Requirements sections into just some text in Future Work.

\makeatletter
\usepackage{enumitem}
\usepackage{multicol}

@@

\subsection{Memory Requirements\label{sub:memoryreq}}
\remi{maybe its own ``limitations'' section? also about
  static segmentation and limited scaling }

There are several sources of extra memory requirements in our
STM system. First, we need to keep track of the state an object
is in. We do this using flags and an overflow number. Both currently
fit in a single additional header of 4~bytes per object.

Second, there are areas in memory private to each segment (see
Section~\ref{sub:Setup}). The Nursery for example is 4~MiB in
size\footnote{Configurable value taken from experience with our
non-STM reference interpreter}. Additionally, the area used for the
read markers is lazily mapped by the kernel and only needed for
objects in the old object space. Thus, it can be at most the total
size of memory used by old objects divided by 16, adding up to
$6.3\%$ to the total memory requirements.

Third, since we do copy on write, if all objects were to be written to
at the same time, all pages would need to be privatised for all
objects in the old object space. In that case we would need the total
amount of memory required by old objects multiplied by $N+1$
(incl. the sharing segment). This is certainly the worst case and
so far not observed in practice. During major collections we re-share
privatised pages that are currently not modified by any transaction,
thereby reclaiming some of this memory.

@@

\subsection{Memory Requirements}

\remi{reference \ref{sub:memoryreq}}

% \begin{figure}[h]
%   \centering
%   \includegraphics[width=1\columnwidth]{plots/richards_mem.pdf}
%   \caption{Actual memory managed by the GC and the page privatisation
%     over time in Richards benchmark\label{fig:richards_mem}}
% \end{figure}

% In Figure~\ref{fig:richards_mem} we look at the memory usage of one of
% our benchmarks called Richards.\footnote{Richards is a OS kernel simulation
% benchmark.} The \emph{GC managed memory} counts all memory used in the
% old object space including the memory required for private pages. Note,
% it does not measure the memory required for the segment-private areas
% like the nursery and read markers. The
% sharp drops in memory usage come from major collections that free old
% objects and re-share pages. The average memory usage stays around
% 29~MiB and we see that in this benchmark we have around 1 major
% collection every second.

% The \emph{page privatisation}, which represents the ratio between
% private pages and shared pages, gives us an idea about how many
% private copies exist for each shared page. If the ratio is $>1.0$, it
% means that on average we have more than one private copy for each
% shared page. In the worst case described above, this number would
% reach $N$ and means we need $N\times$ the memory for the private
% pages alone. In this benchmark, we see the same spikes as the memory
% usage. These come directly from re-sharing the pages. The maximum page
% privatisation is around $0.9$ between major collections, thus the
% private pages are responsible for a $90\%$ increase in the required
% memory. Since the spikes in the GC managed memory line actually
% show increases by $~100\%$, it means that the rest comes from actual
% garbage objects that were collected.

\remi{do this and RSS numbers for pypy-stm-nojit and normal pypy-nojit:}

For PyPy-STM the average memory requirement is 21~MiB and there are
$\sim 10$ major collections during the runtime. Normal PyPy with a GIL
grows its memory up to just 7~MiB and does not do a single major
collection in that time. Compared to normal PyPy, we are missing a
memory optimisation to store small objects in a more compact
way. Additionally, since normal PyPy uses a GIL, it does not need to
duplicate any data structures like e.g.\ the Nursery for each
thread. This, the missing optimisation, and the additional memory
requirements for STM explained above account for this difference.
We expect to improve this aspect in the future; in this paper we
want to focus first on performance.
\remi{I don't know how much sense it makes to go deeper. We will
improve this in the future, but right now this is the overall picture.}
\cfbolz{you can't kill it, otherwise a reviewer will ask. also, the pypy-no-stm numbers are meaningless as long as the options are different}

@@

>>>>

\makeatletter
\usepackage{enumitem}
\usepackage{multicol}
\usepackage{tabularx}

@@

@@

@@

\section{Future Work}

\arigo{what else than memory usage goes here?}

We are still working on a number of optimizations in the
implementation.  We want to mention notably that memory usage has not
been optimized so far.  There are several sources of extra memory
requirements in our STM system.  The multiple nurseries add
4~MiB\footnote{Configurable value taken from experience with our
  non-STM reference interpreter.} per segment.  Additionally, the area
used for the read markers uses, per segment, up to 1/16th of the
memory used by the old objects.  With four segments, this adds $25\%$
to the total memory requirements.

One of the most important sources of additional memory usage is the
presence of privatised pages.  It is hard to predict and control.
Right now, if a program creates a large number of privatised pages,
major garbage collections will occur more frequently in the hope of
resharing some of these pages.  More advanced techniques might be
useful.

Overall, early measures seem to show that a four-segment version of
pypy-stm-nojit consumes between 2 and 7 times as much memory as a
plain pypy-nojit.  The high part of this spectrum seems to occur only
if the total memory usage is rather low (less than 20~MiB): it might
simply mean that our approach has a high constant memory usage
overhead.  The low part, i.e.\ a factor 2, can approximately be
attributed to the reasons described above.  It is future work to
investigate in more details.

### 2014-06-05 talk/pycon2010/crossinterp/talk.tex
Attempt to simplify the Memory Requirements sections into just some text in Future Work.

@@

\begin{frame}[fragile]
  \frametitle{Don't rely on refcounting}
  \begin{itemize}
    \item example
      \begin{verbatim}
open('x', 'w').write('stuff')
      \end{verbatim}
    \item on refcounting, flushes file immediately
    \item on any other GC, it might be deferred for
      a while
    \item the single most-common problem when fixing twisted
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{\_\_del\_\_}
  \begin{itemize}
    \item resurrection on CPython will call {\ttfamily \_\_del\_\_} multiple times,
      other Pythons exactly once
    \item cycles with \_\_del\_\_s are not collected by CPython,
      PyPy breaks them randomly instead
    \item in PyPy and Jython \_\_del\_\_ cannot be attached to classes after creation
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Use new-style classes}
  \begin{itemize}
    \item 3.x ready
    \item faster on PyPy, too
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{{\ttfamily sys.prefix}}
  \begin{itemize}
    \item implementations may have different installation layouts
    \item open issue, at least for PyPy, has compatibility consequences with
  setuptools
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{IO bytes vs unicode}
  \begin{itemize}
    \item convert/decode as soon as possible, keep text and bytes apart
    \item for 2.x Pythons use str for bytes and unicode for text
    \item the distinction is deeper in 3.x (str is unicode, bytes exist with
  slightly different interface than old str)
  \end{itemize}
\end{frame}

>>>>

\begin{frame}
  \frametitle{My background}
  \begin{itemize}
    \item worked a lot with PyPy compatibility issues
    \item helped to port twisted, django and other projects to run on PyPy
    \item a lot of cooperation with Jython people
  \end{itemize}
\end{frame}

@@

\begin{frame}[fragile]
  \frametitle{Don't rely on refcounting}
  \begin{itemize}
    \item example
      \begin{verbatim}
open('x', 'w').write('stuff')
      \end{verbatim}
    \item on refcounting, flushes file immediately
    \item on any other GC, it might be deferred for
      a while
    \item the single most-common problem when porting twisted to PyPy
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{\_\_del\_\_}
  \begin{itemize}
    \item resurrection on CPython will call {\ttfamily \_\_del\_\_} multiple times,
      other Pythons exactly once
    \item cycles with {\ttfamily \_\_del\_\_}s are not collected by CPython,
      PyPy breaks them randomly instead
    \item in PyPy and Jython {\ttfamily \_\_del\_\_} cannot be attached to classes after creation
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Use new-style classes}
  \begin{itemize}
    \item 3.x ready
    \item much faster on PyPy, too
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{{\ttfamily sys.prefix}}
  XXX think about
  \begin{itemize}
    \item implementations may have different installation layouts
    \item open issue, at least for PyPy, has compatibility consequences with
  setuptools
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{IO bytes vs unicode}
  \begin{itemize}
    \item convert/decode as soon as possible, keep text and bytes apart
    \item for 2.x Pythons use str for bytes and unicode for text
    \item the distinction is deeper in 3.x ({\ttfamily str} is {\ttfamily unicode}, {\ttfamily bytes} exist with
  slightly different interface than old {\ttfamily str})
  \end{itemize}
\end{frame}

\begin{frame}
  \frametitle{Don't concatenate strings}
  \begin{itemize}
    \item use {\ttfamily "".join(...)}
    \item if you care about performance, try this and {\ttfamily cStringIO}
  \end{itemize}
\end{frame}
